---
title: "Modelling Wildlife Corridors: Spatial Analysis of Topographic and
Landscape Barriers to determine Ecological Connectivity" 
subtitle: "Project Work 2"
institut: "INSTITUT FÜR COMPUTATIONAL LIFE SCIENCES "
keywords: "Wildlife Corridors, Connectivity Modeling, Ecological Connectivity, Least-Cost Path (LCP) Analysis, Spatial Analysis"
lang: en
confidential: false
cover-image:
  src: cover.png
  max-width: 100%
  max-height: 5cm
thesis-type: "Projekt Work 2"
degree-type: "Bachelor's degree program"
study-year: "2025"
submission-date: "2025-12-11"
study-direction: "Applied Digital Life Sciences (ADLS)"
supervisors:
  - title: ""
    name: "Ratnaweera Nils"
    affiliation: "ZHAW Life Sciences und Facility Management, Wädenswil"
bibliography: references.bib
author: 
  - Lukas Buchmann
format: 
  zhaw-lsfm-typst: default
  html: default
engine: jupyter
jupyter: python3
---

# Zusammenfassung {.unnumbered .unlisted}

Diese Arbeit untersucht die Beziehung zwischen Gehirngröße und Körpergewicht bei verschiedenen Säugetierarten anhand des MASS::Animals Datensatzes. Die Analyse zeigt eine starke positive Korrelation zwischen beiden Variablen, wobei sich jedoch erhebliche Abweichungen von der erwarteten allometrischen Beziehung ergeben. Besonders Primaten zeigen ein überproportional grosses Gehirn-Körper-Verhältnis. Die Ergebnisse bestätigen frühere Erkenntnisse zur Evolution der Gehirngröße und liefern wichtige Einblicke in die Allometrie des Nervensystems.


::: {.content-hidden when-format="html"}

# Inhaltsverzeichnis {.unnumbered .unlisted}

{{< table-of-contents >}}

:::



# Liste der Abkürzungen {.unnumbered}

+---------+----------------------------------+
| MASS    | Modern Applied Statistics with S |
+---------+----------------------------------+
| log     | natürlicher Logarithmus          |
+---------+----------------------------------+
| R²      | Bestimmtheitsmass                |
+---------+----------------------------------+
| kg      | Kilogramm                        |
+---------+----------------------------------+
| g       | Gramm                            |
+---------+----------------------------------+  

# Introduction

## 1.1 The Challenge of Habitat Fragmentation
Across Europe, landscapes are undergoing rapid transformation driven by the expansion of transportation infrastructure, urban sprawl, and the intensification of agriculture (European Environment Agency, 2024). This process leads to habitat fragmentation: the division of large, continuous habitats into smaller, more isolated patches. Fragmentation is recognized as a primary driver of global biodiversity loss (Haddad et al., 2015). Its ecological consequences are severe: it limits wildlife mobility, isolates populations, and restricts gene flow (Kuehn et al., 2007; Wang et al., 2016). This, in turn, can lead to inbreeding, a loss of adaptive genetic diversity, and reduced long-term population resilience, making populations more vulnerable to local extinction (Haddad et al., 2015; Kuehn et al., 2007).

The situation in Switzerland is particularly acute (Jaeger, 2007; Jaeger et al., 2011). The Swiss Federal Office for the Environment (FOEN) has identified fragmentation from artificial barriers as a key pressure on biodiversity (FOEN, 2010; Jaeger et al., 2011). This fragmentation is most severe in the Swiss Central Lowlands (FOEN, 2010). A comprehensive study by the Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) identified the Canton of Schaffhausen as one of the eight most fragmented cantons in the country, underscoring the urgent need for regional connectivity planning (Jaeger, 2007).

## 1.2 The Role of Ecological Connectivity
To counteract these effects, conservation efforts increasingly focus on maintaining and restoring ecological connectivity, defined as the unimpeded movement of species and the flow of natural processes that sustain life. A primary tool for achieving this is the identification and protection of wildlife corridors. These are linear landscape elements that link otherwise isolated habitat patches. By facilitating animal movement, corridors are essential for enabling dispersal, maintaining genetic diversity, and allowing species to migrate in response to seasonal needs or long-term climate change (Hilty et al., 2020).

## 1.3 Focal Species: The Roe Deer (Capreolus capreolus)
The successful planning of such corridors depends on a species-specific approach (Beier et al., 2008). This project focuses on the roe deer (Capreolus capreolus), an ideal focal species for connectivity modelling in the Swiss landscape. As one of Europe's most common wild ungulates, the roe deer is highly adaptable. It demonstrates significant behavioral flexibility, inhabiting not only its traditional forest-mosaic habitats but also does well in open agricultural plains (Jepsen et al., 2012).

Despite this adaptability, roe deer are highly vulnerable to fragmentation. Transportation networks pose a dual threat: direct mortality from wildlife-vehicle collisions and the barrier effect, where high-traffic roads hinder access to critical resources (Maertz et al., 2024). This barrier effect can be as damaging as direct habitat loss, leading to the genetic isolation of populations. A study on roe deer in Central Switzerland, for example, demonstrated significant genetic differentiation between populations separated by a fenced motorway. Because roe deer are widespread, mobile, and frequently interact with infrastructure, they serve as an excellent model for assessing landscape-level connectivity for large mammals (Kuehn et al., 2007).

## 1.4 Methodology: Least-Cost Path (LCP) Analysis
This study employs Least-Cost Path (LCP) analysis, a standard and widely used methodology in GIS-based connectivity modelling (Beier et al., 2008; Cushman et al., 2013). The LCP approach is built on the concept of a resistance surface (Cushman et al., 2013). This is a GIS raster layer where each pixel is assigned a "cost" value representing the difficulty, energy expenditure, or mortality risk a species encounters when moving through that specific landscape type (Zeller et al., 2012). This "cost" is derived from expert knowledge and ecological literature, assigning low resistance to preferred habitats and high resistance to barriers (Beier et al., 2008). The LCP algorithm then calculates the most efficient route between defined core habitat patches, the path of least cumulative resistance (Cushman et al., 2013).

## 1.5 Project Objectives
The primary objective of this project is to develop, improve, and evaluate a GIS-based connectivity model for roe deer (Capreolus capreolus) in the highly fragmented landscape of the Canton of Schaffhausen, implemented using the Python programming language.

The specific aims are to:

- Create a detailed resistance surface using OpenStreetMap and Corine Land Cover data.

- Perform a Least-Cost Path (LCP) analysis to compute cumulative movement costs and identify potential wildlife corridors connecting suitable habitat patches.

- Analyze the resulting model to identify and map key bottlenecks, major obstacles, and other landscape patterns that limit ecological connectivity.

The final output will be a reproducible connectivity model and a set of cartographic products. This work is intended to serve as a practical, data-driven decision-making aid for regional conservation and spatial planning.

# Theoretical Framework

## Ecological Connectivity and Wildlife Corridors
### Functional vs. Structural Connectivity
There are two distinct forms of Landscape connectivity: structural and functional. Structural connectivity refers only to the physical arrangement and contiguity of landscape elements, such as the spatial proximity of habitat patches, without regard for the specific mobility of an organism. In contrast, functional connectivity explicitly includes the behavioral response of a species to these landscape features and describes the degree to which the landscape actually facilitates or impedes movement (Taylor et al., 1993; Tischendorf & Fahrig, 2000).

Although structural connectivity can be measured using landscape metrics alone, this often fails to predict the actual movement of organisms when the matrix between habitats is hostile. For example, two forest patches might be structurally close, but functionally isolated by a highway. That is why this project focuses on a functional approach. By integrating the roe deer's specific movement behavior and habitat preferences into the model, the resulting connectivity map reflects the organism's ability to traverse the landscape, rather than merely the geometric arrangement of features (Haddad et al., 2015).

### Types of Corridors
In the context of connectivity conservation, corridors are generally classified into two functional categories: continuous linear corridors and stepping stones.

Linear corridors are unbroken strips of suitable habitat that physically connect two larger core areas. Typical examples include hedgerows, riparian strips along rivers, or wooded belts. For ungulates like the roe deer, such continuous features are vital as they offer uninterrupted cover and forage during movement, significantly reducing predation risk and stress.

Stepping stones, in contrast, are a series of small, discontinuous habitat patches located within a hostile matrix. While they do not provide a physical connection, they facilitate movement by offering temporary refuge for resting and feeding during Movement (Saura et al., 2014). In fragmented cultural landscapes like the Swiss Plateau, where continuous forest belts are often interrupted by agriculture and infrastructure, these stepping stones are essential for maintaining functional connectivity(Jaeger et al., 2008; Saura et al., 2014).

For this GIS analysis, the distinction is crucial: while the Least-Cost Path (LCP) algorithm seeks a continuous route of minimized resistance, the biological reality of the roe deer allows it to traverse short distances of higher-resistance "matrix" (e.g., open fields) to reach the next stepping stone. Consequently, the resistance surface must be weighted to penalize open areas and barriers without rendering them completely impassable, thereby modeling the landscape's overall permeability rather than just identifying structural links.

## Roe Deer (Capreolus capreolus) Ecology
### Habitat Preferences
Ecologically, the roe deer (Capreolus capreolus) is classified as a "concentrate selector" (Hofmann, 1989). Unlike grazers that can digest high-fiber grass, roe deer require high-energy, easily digestible plant parts such as buds, shoots, and herbs. This physiological constraint dictates a close association with habitats offering high plant diversity (Holand et al., 1998).

Roe deer are characterized as "edge species," exhibiting a strong preference for the transition zones (ecotones) between closed forest cover and open agricultural land. These zones allow them to optimize the trade-off between accessing high-quality forage in open fields and maintaining immediate access to cover for predator avoidance (Tufto et al., 1996; Mysterud & Østbye, 1999). Consequently, forest environments (landuse=forest, natural=wood) represent the primary diurnal resting habitat and are essential for concealment. Open habitats such as meadows, orchards, and farmlands are utilized primarily during nocturnal or crepuscular periods for foraging, but lack the necessary cover for permanent habitation.

### Movement capabilities and Physical Constraints
While roe deer are physically capable of swimming, they generally avoid crossing large water bodies **because these areas offer no cover, leaving the animal exposed**. Consequently, large rivers and lakes act as significant barriers, whereas smaller streams are passable but associated with higher movement costs due to the lack of concealment. Similarly, wetlands (natural=wetland) are physically passable but energetically costly due to unstable ground conditions and lack of dense canopy cover compared to forest environments.

### Barriers and Avoidance
Anthropogenic features constitute the primary barriers to connectivity. Roe deer are highly sensitive to visual and acoustic disturbances:

- Urbanization: Settlements and industrial areas are generally avoided due to human presence and lack of vegetation.

- Infrastructure: Heavily used roads and railways act as barriers not only due to the physical risk of collision but also due to traffic noise, which deters crossing (Kuehn et al., 2007).

- Linear Obstacles: Fences and dense hedges can physically impede movement, effectively fragmenting habitat patches.

- Recreation: Small paths are structurally passable, but human recreational activity—specifically the presence of walkers and dogs during the day—creates a "landscape of fear," causing deer to avoid these areas despite low physical resistance.

## Least-Cost Path (LCP) Analysis
### The Resistance Surface
The resistance surface is the fundamental spatial data layer in connectivity modeling. It is a raster grid where every cell (pixel) is assigned a numerical value representing the "cost" or "energy" an organism needs when traversing that specific location (Zeller et al., 2012). This value integrates the biological factors defined in Section 2.2 and combines land cover suitability, anthropogenic barriers, and topographic constraints into a single indicator.

Mathematically, the resistance surface represents the reciprocal of landscape permeability. High values indicate barriers or hostile matrix (high energy expenditure or mortality risk), while low values indicate preferred habitat (low energy expenditure). The accuracy of the entire LCP analysis is contingent upon the biological validity of these resistance assignments (Beier et al., 2008).

### Cost Distance Calculation
Unlike Euclidean distance, which measures the "as-the-crow-flies" separation between two points, cost distance measures the cumulative "expense" of movement. The calculation employs graph theory algorithms, most notably a raster-based adaptation of Dijkstra’s algorithm (Dijkstra, 1959; Adriaensen et al., 2003).

The algorithm treats the center of each raster cell as a node and connects it to its neighboring cells (usually the 8 immediate neighbors). The cost to move between two adjacent cells is calculated as the distance multiplied by the average resistance of the two cells.

The algorithm propagates outwards from the source habitat, calculating the "Accumulated Cost" to reach every other pixel in the study area. The result is an Accumulated Cost Surface, where the value of a pixel represents the lowest possible total cost to travel to that location from the source, avoiding high-resistance barriers where possible (Etherington, 2016).

### The Path
The final Least-Cost Path is derived from the Accumulated Cost Surface. The algorithm identifies the Target node (the destination habitat) and "backtracks" to the Source node by consistently choosing the steepest downhill gradient on the accumulated cost surface.

Ecologically, this single line represents the theoretically optimal route where an animal minimizes energy expenditure and risk (Sawyer et al., 2011). However, it is important to note that LCP models assume the animal possesses "omniscience" (perfect knowledge) of the landscape, which is a theoretical idealization. In reality, animals may deviate from this optimal path. Therefore, the LCP should be interpreted as the center line of the most probable movement corridor rather than a deterministic route (Cushman et al., 2013).

# Material and Methods

## Study Area
The study focuses on the Canton of Schaffhausen, situated in Northern Switzerland, as defined in the project disposition. Geographically, the region is characterized by a heterogeneous mosaic of scrub and mixed forests alternating with intensively farmed agricultural land and expanding settlements. The area is dissected by the Rhine River to the south and a dense network of transport infrastructure throughout.

This region was selected specifically because it has been identified as one of the most fragmented cantons in Switzerland (Jaeger, 2007). Consequently, it serves as a critical case study for modeling connectivity in a pressured Central European landscape.

To ensure geometric consistency across the transboundary study area (Switzerland and Germany), all spatial datasets were reprojected to EPSG:32632 (WGS 84 / UTM zone 32N). The analysis extent covers the political boundaries of the canton, including a buffer zone of 1000m to minimize edge effects during the Least-Cost Path calculation.

(See Figure 1: Map of the Study Area Schaffhausen showing major land cover types)

## Software and Computational Environment
The analysis was conducted using Python 3.10 within a reproducible environment. The computational workflow consisted of three distinct phases: (1) the generation of the resistance surface, (2) the calculation of least-cost paths between core habitats, and (3) the statistical analysis and visualization of the resulting corridors.

The key open-source Python libraries employed for geospatial analysis included:

GeoPandas **(v. 1.0+)**: Used for vector data manipulation, including clipping and buffering of study area boundaries.

Rasterio **(v. 1.3+)**: Employed for all raster operations, specifically reprojection, rasterization, and the weighted overlay of resistance layers.

scikit-image **(v. 0.2x)**: Used to execute the Least-Cost Path (LCP) algorithm via the MCP_Geometric class.

## Data Preprocessing and Preparation

### Geospatial Datasets
To ensure reproducibility and accessibility, the model relies exclusively on open-source geospatial data.

#### Primary Feature Data: OpenStreetMap (OSM) 
High-resolution landscape features were derived from OpenStreetMap (OSM). This Volunteered Geographic Information (VGI) dataset provides the fine-scale geometric detail required to model barriers such as fences, minor roads, and buildings, which are crucial for wildlife movement analysis.

Acquisition: Raw data was retrieved in Protocolbuffer Binary Format (.pbf) from Geofabrik GmbH (Geofabrik, 2025). To cover the transboundary study area, extracts for Switzerland and the administrative region of Baden-Württemberg are automatically downloaded.

Processing: The data was extracted and clipped to the Area of Interest (AOI) using the pyrosm library. The dataset was filtered to retain only the ecologically relevant features in osm_resistance_costs.csv.

Projection: All vectors were reprojected to the project’s master CRS, EPSG:32632 (UTM 32N), to ensure metric accuracy in distance calculations.

#### Base Land Cover Data: CORINE Land Cover (CLC) 
The CORINE Land Cover (CLC) 2018 dataset served as the continuous background layer.

Description: CLC provides standardized European land cover classes at a 100-meter resolution with a 25-hectare minimum mapping unit.

Acquisition: The raster dataset (U2018_CLC2018_V2020_20u1.tif) was acquired from the Copernicus Land Monitoring Service (Copernicus, 2020).

Role in Model: Due to its coarser resolution, CLC acts as a "fallback layer." It assigns a generalized resistance value to any pixel where fine-scale OSM data is absent, ensuring the final cost surface is continuous and free of "NoData" gaps.

### Derivation of the Resistance Surface
A hybrid rasterization approach was developed to create the final cost surface. To ensure transparency and reproducibility, all resistance values were defined in two external configuration files (osm_resistance_costs.csv and clc_resistance_costs.csv).

The model operates on a "Burn-In" principle: High-resolution OSM features are rasterized on top of the CLC base map. Resistance values range from 1 (optimal habitat) to 5000 (absolute barrier).

#### The Base Layer (clc_resistance_costs.csv) 
This configuration file maps all 44 CLC Level-3 classes to ecological resistance values based on Roe Deer habitat preferences:

- Optimal Habitat (Cost 1): Broad-leaved forest (311) and Mixed forest (313).

- Permeable Matrix (Cost 8–20): Pastures (231) and Non-irrigated arable land (211).

- Barriers (Cost 800–5000): Water bodies (512) and Continuous urban fabric (111).

#### The Feature Layer (osm_resistance_costs.csv) 
This file controls the rasterization of OSM features. It contains four columns: osm_key, osm_value, resistance, and priority.

Ecological Cost (resistance):

- Transportation: **Costs follow the COST 341 framework guidelines**, creating a gradient from highway=track (Cost 20) to fenced highway=motorway (Cost 5000).

- Barriers: Explicit obstacles such as barrier=fence (Cost 800) and barrier=wall (Cost 1500) are included.

Technical Implementation (priority): The priority column dictates the rasterization order (burn-in sequence). This ensures correct layering. For example, a highway (Priority 10) is rasterized after a forest polygon (Priority 1), ensuring the road "cuts through" the forest in the final model.

### Vector to Raster Conversion
To integrate the vector-based Data into the grid-based resistance model, a rasterization process was executed using the features.rasterize module from the Rasterio library (Gillies, 2013).

A critical technical specification in this workflow was the activation of the all_touched=True parameter. Standard rasterization algorithms typically use a "center-point" rule, where a pixel is only assigned a value if the vector geometry covers its center. This method frequently results in "rasterization artifacts," where thin or diagonal linear features appear as broken, discontinuous chains of pixels.

By enforcing the all_touched=True rule, any pixel intersected by a vector geometry was assigned the feature's resistance value. This step was essential to ensure the topological continuity of linear barriers (e.g., fences, walls, and narrow streams). Without this parameter, the rasterization process would create artificial gaps in these barriers, through which the Least-Cost Path algorithm could incorrectly route a wildlife corridor, leading to biologically invalid results.

### Reclassification and Standardization

## Development of the Resistance Surface
### Weighting Scheme
To translate the ecological behaviors into the resistance model, the specific OpenStreetMap (OSM) tags were assigned resistance values ranging from 1 (optimal habitat) to 5000 (severe barrier).

Methodological Note on Absolute Barriers: In this model, no landscape feature (including buildings or highways) was assigned a value of "NoData" or "Infinity." Instead, they were assigned very high, but finite, resistance values (e.g., 5000). This assumption is necessary to identify bottlenecks. If the Least-Cost Path (LCP) is forced to traverse a building, it indicates that the surrounding landscape is so impermeable that no better alternative exists. This highlights areas where connectivity is critically compromised and restoration (e.g., removing a barrier) is required.

Table 3.1: Justification of Resistance Values
OSM Key,OSM Value,Resistance,Justification for Roe Deer Suitability,Source
Optimal Habitat (Cover),,,,
landuse / natural,"forest, wood",1,Primary habitat. Provides optimal cover and protection; essential for resting during the day.,[Hab-Ziolkowska]
natural,scrub,3,"Good cover, but slightly less navigable or protective than mature forest.",[Hab-Ziolkowska]
natural,heath,8,"Provides some low cover and forage, but less protection than forest/scrub.",[Hab-Ziolkowska]
Foraging Habitat (Open),,,,
landuse / natural,"meadow, grassland, grass",5,"Important foraging resources. Low resistance for movement at night, but lacks cover (higher predation risk than forest).",[Hab-Ziolkowska]
landuse,orchard,10,"Good food source (fruit/buds) with moderate cover from trees, but often fenced or managed.",[Hab-Ziolkowska]
landuse,vineyard,15,"Provides food, but typically terraced or fenced, making movement physically more demanding.",[Hab-Ziolkowska]
landuse,farmland,20,High food availability but very high exposure risk; often lacks cover for long distances.,[Hab-Ziolkowska]
barrier,hedge,30,"Can serve as a linear corridor (cover), but dense hedging can also act as a physical barrier to cross.",[Hab-Ziolkowska]
Disturbed / Recreational,,,,
highway,"path, footway, cycleway, bridleway, track",10 - 20,"Physically easy to walk on, but avoided during the day due to human and dog presence (disturbance).",[PA2-Logic]
landuse,allotments,100,Mix of food and fences; high human presence and physical obstacles.,[Hab-Ziolkowska]
landuse / leisure,"recreation_ground, park, cemetery",150 - 250,Open areas within settlements; high human disturbance makes them poor corridors despite vegetation.,[Hab-Ziolkowska]
leisure,golf_course,300,"Highly managed, open, often fenced; high human activity.",[Hab-Ziolkowska]
Natural Barriers,,,,
waterway,ditch,50,Minor physical obstacle; requires jumping.,[PA2-Logic]
natural,wetland,75,Passable but physically exhausting ground conditions; lack of canopy cover.,[Expert-Comp]
waterway,stream,150,Physical barrier; lack of cover while crossing increases stress.,[Expert-Comp]
natural,"scree, bare_rock",250 - 300,Difficult terrain; lack of food and cover; risk of injury.,[Expert-Comp]
natural,cliff,4000,Severe physical barrier; nearly impossible to traverse.,[PA2-Logic]
waterway,river,4000,Major barrier. Requires swimming (high exposure/risk); only crossed in emergencies.,[Expert-Comp]
Anthropogenic Barriers (Infrastructure),,,,
highway,"residential, service, unclassified",200 - 250,"Low traffic speed, but constant presence of cars and people.",[Infra-COST341]
highway,"tertiary, tertiary_link",500,Moderate traffic noise and volume; acts as a behavioral deterrent.,[Infra-COST341]
barrier,fence,800,"Physical obstacle; permeability depends on height, but modeled as high resistance.",[PA2-Logic]
highway,secondary,1000,High traffic volume and noise; significant mortality risk and behavioral avoidance.,[Infra-COST341]
landuse,quarry,1500,Hazardous terrain with industrial disturbance.,[Expert-Comp]
barrier,wall,1500,Solid physical barrier; impassable unless breached.,[PA2-Logic]
highway,primary,2000,Very high traffic/noise. Crossing is rare and highly dangerous.,[Infra-COST341]
railway,"rail, landuse=railway",3000,Noise from trains is a major deterrent; tracks and embankments are physical hurdles.,[Infra-COST341]
highway,"motorway, trunk",5000,Maximum Infrastructure Barrier. Fencing and high-speed traffic make crossing nearly impossible.,[Infra-COST341]
Anthropogenic Barriers (Settlement),,,,
landuse,residential,3500,"High density of buildings, fences, and people. Only traversed at edges.",[Expert-Comp]
landuse,"commercial, industrial, retail",4500,"Large sealed surfaces, high noise, fences, and truck traffic. Very hostile.",[Expert-Comp]
building,yes,5000,Theoretical Maximum Cost. Physically impassable. Included as valid data to identify critical conflicts if the LCP is forced through a structure.,[Expert-Comp]

### Integration of Topographic Variables
The creation of the resistance surface was the foundational step for this project, and it was built as a modular, reproducible workflow using Python, primarily with the GeoPandas, Rasterio, and NumPy libraries. Three documents are required to execute the workflow. The `resistance_costs.csv` file, which stores the resistance costs for each landcover, water and roads type, the `gis_utils.py` file, which containes all gis, loading and plotting functions for processing the data and the `cost_surface_model.ipynb` file, which executes the workflow step by step.

## Connectivity Modelling
The connectivity analysis was performed using a node-based approach, where "nodes" represent core habitat patches capable of supporting a roe deer population. The modeling process involved two distinct stages: (1) the identification of suitable source and target patches, and (2) the calculation of the least-cost paths connecting them.




::: {.content-visible when-format="html"}

::: {.content-visible when-format="html"}

```{python}
#| label: fig-map-interactive
#| fig-cap: "Interaktive Analyse der Widerstandsflächen"
#| echo: false
#| warning: false
#| output: true
#| cache: true

import glob
import os
import xarray as xr
import rioxarray
import hvplot.xarray
import holoviews as hv

hv.extension('bokeh')

input_dir = r"C:\ZHAW\5.Semester\PA2\PA2-Modelling_Wildlife_Corridors\Results\temp_files"
tif_files = sorted(glob.glob(os.path.join(input_dir, "*.tif")))

if tif_files:
    # 1. Globale Min/Max Berechnung (auf stark verkleinerten Daten)
    global_min = float('inf')
    global_max = float('-inf')
    
    # Wir lesen nur eine sehr grobe Übersicht für die Statistik
    for f in tif_files:
        # overview_level=3 bedeutet ca. 1/8 der Auflösung -> Extrem schnell
        with rioxarray.open_rasterio(f, overview_level=3, masked=True) as da_stat:
             mn = da_stat.min().item()
             mx = da_stat.max().item()
             if mn < global_min: global_min = mn
             if mx > global_max: global_max = mx

    map_overlay = None
    
    # 2. Karten-Erstellung
    for f in tif_files:
        # Laden mit overview_level=1 (halbe/viertel Auflösung) ist oft genug für Web
        # Das spart das teure .coarsen() und das Laden der vollen 45MB!
        da = rioxarray.open_rasterio(f, overview_level=1, masked=True)
        
        # Reprojezieren (immer noch nötig, aber jetzt viel schneller)
        if da.rio.crs != "EPSG:3857":
             da = da.rio.reproject("EPSG:3857")
        
        layer_name = os.path.basename(f)
        
        p = da.hvplot.image(
            x='x', y='y', 
            rasterize=True, 
            cmap='RdYlGn_r', 
            clim=(global_min, global_max),
            alpha=0.7, 
            tiles='OSM', 
            name=layer_name,
            frame_width=700, frame_height=500,
            colorbar=True
        )
        
        if map_overlay is None:
            map_overlay = p
        else:
            map_overlay = map_overlay * p

    map_overlay
```
:::

::: {.content-visible when-format="typst"}

```{python}
#| label: fig-map-static
#| fig-cap: "Final Resistance Surface (Statische Ansicht)"
#| echo: false
#| warning: false

import matplotlib.pyplot as plt
import rioxarray
import os

# Hier nehmen wir nur das eine finale File für das PDF
static_file = r"C:\ZHAW\5.Semester\PA2\PA2-Modelling_Wildlife_Corridors\Results\final_resistance_surface.tif"

if os.path.exists(static_file):
    # Datei laden
    da = rioxarray.open_rasterio(static_file, masked=True).squeeze()
    
    # Plotten mit Matplotlib
    fig, ax = plt.subplots(figsize=(10, 8))
    da.plot(ax=ax, cmap='RdYlGn_r', add_colorbar=True, cbar_kwargs={'label': 'Resistance Value'})
    
    ax.set_title("Final Resistance Surface")
    ax.set_axis_off() # Koordinatenachsen ausblenden für saubereren Look
    
    plt.tight_layout()
    plt.show()
else:
    print(f"Datei nicht gefunden: {static_file}")
```
:::

### Transformationsverfahren

Die logarithmische Transformation ist bei allometrischen Beziehungen Standard, da sie Potenzgesetze in lineare Beziehungen überführt. Dies ermöglicht die Anwendung klassischer linearer Regressionsverfahren.

### Modellvalidierung

Die Residuen wurden auf Normalität und Homoskedastizität überprüft. Zusätzlich wurden Cook's Distance und Leverage-Werte berechnet, um einflussreiche Datenpunkte zu identifizieren.

## Visualisierung

Scatterplots wurden erstellt, um die Beziehung zu visualisieren, sowohl auf ursprünglicher als auch auf logarithmischer Skala.

### Graphische Darstellung

Die Visualisierungen umfassen sowohl die Rohdaten als auch die transformierten Daten mit Regressionsgerade und Konfidenzintervall.

### Software und Pakete

Alle Analysen und Visualisierungen wurden in R durchgeführt unter Verwendung der Pakete ggplot2, MASS und dplyr.

# Ergebnisse

Die statistische Analyse des MASS::Animals Datensatzes liefert umfassende Einblicke in die Beziehung zwischen Körper- und Gehirngewicht bei Säugetieren. Die deskriptiven Statistiken in @tbl-summary zeigen die erhebliche Variabilität der untersuchten Arten: Das Körpergewicht reicht von kleinen Nagern bis hin zu grossen Säugetieren wie Elefanten, während das Gehirngewicht entsprechend von wenigen Gramm bis zu mehreren Kilogramm variiert.

Die logarithmische Darstellung in @fig-animals visualisiert deutlich die allometrische Beziehung zwischen beiden Variablen. Trotz der logarithmischen Transformation zeigen sich einige interessante Abweichungen von der Regressionsgerade, die auf artspezifische Anpassungen hinweisen. Besonders auffällig sind Datenpunkte oberhalb der Regressionslinie, die Arten mit überproportional grossem Gehirn repräsentieren.

```{r}
#| label: fig-animals
#| fig-cap: "Beziehung zwischen Körpergewicht und Gehirngewicht bei Säugetieren"

library(MASS)
library(ggplot2)

data(Animals)

# Scatterplot auf logarithmischer Skala
ggplot(Animals, aes(x = body, y = brain)) +
  geom_point(size = 3, alpha = 0.7, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "Körpergewicht (kg)", 
    y = "Gehirngewicht (g)",
    title = "Gehirn- vs. Körpergewicht (logarithmische Skala)"
  ) +
  theme_minimal()
```

```{r}
#| label: tbl-animals
#| tbl-cap: "Statistische Kennwerte der Gehirn-Körper-Beziehung"

# Regressionsanalyse
model <- lm(log(brain) ~ log(body), data = Animals)
summary_stats <- summary(model)

# Tabelle erstellen
library(knitr)
coef_table <- data.frame(
  Parameter = c("Intercept (log a)", "Steigung (b)", "R²", "p-Wert"),
  Wert = c(
    round(coef(model)[1], 3),
    round(coef(model)[2], 3),
    round(summary_stats$r.squared, 3),
    ifelse(summary_stats$coefficients[2,4] < 0.001, "< 0.001", 
           round(summary_stats$coefficients[2,4], 3))
  )
)

kable(coef_table)
```

```{r}
#| label: tbl-summary
#| tbl-cap: "Deskriptive Statistik der MASS::Animals Daten"

library(knitr)
summary_table <- data.frame(
  Variable = c("Körpergewicht (kg)", "Gehirngewicht (g)"),
  Minimum = c(min(Animals$body), min(Animals$brain)),
  Mittelwert = c(round(mean(Animals$body), 1), round(mean(Animals$brain), 1)),
  Maximum = c(max(Animals$body), max(Animals$brain)),
  N = c(nrow(Animals), nrow(Animals))
)

kable(summary_table)
```

Die Regressionsanalyse bestätigt die theoretischen Erwartungen zur allometrischen Skalierung. Die statistischen Kennwerte in @tbl-animals zeigen einen Steigungsparameter von 0.75, der optimal in den von der Literatur vorhergesagten Bereich zwischen 0.6 und 0.8 fällt. Das hohe Bestimmtheitsmass (R² = 0.96) belegt die starke lineare Beziehung zwischen logarithmiertem Körper- und Gehirngewicht.

Der hochsignifikante p-Wert (< 0.001) in @tbl-animals unterstreicht die statistische Robustheit dieser Beziehung. Diese Ergebnisse stehen im Einklang mit der allometrischen Theorie und bestätigen, dass das Gehirngewicht systematisch mit dem Körpergewicht skaliert, jedoch nicht proportional, sondern mit einem reduzierten Exponenten.

# Diskussion

Die Ergebnisse bestätigen die grundlegende allometrische Beziehung zwischen Gehirn- und Körpergewicht bei Säugetieren. Der ermittelte Exponent von 0.75 liegt im erwarteten Bereich und stimmt mit früheren Studien überein [@jerison1973].

Besonders interessant sind die Ausreisser in den Daten: Arten wie der Mensch und andere Primaten zeigen ein überproportional grosses Gehirn, was auf spezielle evolutionäre Anpassungen hinweist. Diese Abweichungen unterstreichen die Bedeutung der Encephalisation in der Primatenevolution.

Die hohe Korrelation (R² = 0.96) zeigt, dass das Körpergewicht ein starker Prädiktor für das Gehirngewicht ist, auch wenn bedeutsame biologische Variationen existieren.

# Literaturverzeichnis {#sec-refs}

{{< references >}}


::: {.content-hidden when-format="html"}

# Verzeichnis der Abbildungen {.unnumbered}

{{< list-of-figures >}}

# Verzeichnis der Tabellen {.unnumbered}

{{< list-of-tables >}}

:::

# Anhang {.unnumbered}

Zusätzliche Materialien und Daten, die nicht im Hauptteil der Arbeit enthalten sind, werden im Anhang aufgeführt.

## Rohdaten

Der vollständige MASS::Animals Datensatz mit allen 28 Säugetierarten:

```{r}
#| label: tbl-full-data
#| tbl-cap: "Vollständige Rohdaten des MASS::Animals Datensatzes"

library(MASS)
library(knitr)
data(Animals)

# Add row names as a column and format nicely
animals_full <- data.frame(
  Art = rownames(Animals),
  Körpergewicht_kg = Animals$body,
  Gehirngewicht_g = Animals$brain,
  Log_Körpergewicht = round(log10(Animals$body), 3),
  Log_Gehirngewicht = round(log10(Animals$brain), 3)
)

kable(animals_full, row.names = FALSE)
```

### Datenquelle und -qualität

Die Daten stammen aus verschiedenen Quellen und wurden über mehrere Jahrzehnte zusammengetragen. Einige Besonderheiten der Datensammlung:

- **Afrikanischer Elefant**: Mit 6654 kg Körpergewicht und 5712 g Gehirngewicht das grösste Säugetier in der Stichprobe
- **Spitzmaus**: Das kleinste Säugetier mit nur 0.004 kg Körpergewicht und 0.14 g Gehirngewicht
- **Mensch**: Zeigt mit 1320 g Gehirngewicht bei 62 kg Körpergewicht eine deutliche positive Abweichung von der allometrischen Beziehung
- **Primaten generell**: Schimpanse, Gorilla und Rhesusaffe zeigen alle überdurchschnittlich grosse Gehirne

### Ausreisser-Analyse

```{r}
#| label: fig-residuals
#| fig-cap: "Residuen der logarithmischen Regression zur Identifikation von Ausreissern"

model <- lm(log(brain) ~ log(body), data = Animals)
residuals_df <- data.frame(
  Art = rownames(Animals),
  Fitted = fitted(model),
  Residuals = residuals(model),
  Body = Animals$body
)

library(ggplot2)
ggplot(residuals_df, aes(x = Fitted, y = Residuals, label = Art)) +
  geom_point(size = 3, alpha = 0.7, color = "darkblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_text(aes(label = ifelse(abs(Residuals) > 0.5, Art, "")), 
            vjust = -0.5, hjust = 0.5, size = 3) +
  labs(
    x = "Angepasste Werte (log scale)",
    y = "Residuen",
    title = "Residuen vs. Angepasste Werte"
  ) +
  theme_minimal()
```

## Statistische Outputs

### Vollständige Regressionsanalyse

```{r}
#| label: tbl-full-regression
#| tbl-cap: "Detaillierte Regressionsstatistiken"

model <- lm(log(brain) ~ log(body), data = Animals)
summary_full <- summary(model)

# Create comprehensive statistics table
regression_stats <- data.frame(
  Statistik = c("Multiple R-squared", "Adjusted R-squared", "F-statistic", 
                "F p-value", "Residual standard error", "Degrees of freedom"),
  Wert = c(
    round(summary_full$r.squared, 4),
    round(summary_full$adj.r.squared, 4),
    round(summary_full$fstatistic[1], 2),
    ifelse(pf(summary_full$fstatistic[1], summary_full$fstatistic[2], 
             summary_full$fstatistic[3], lower.tail = FALSE) < 0.001, 
           "< 0.001", 
           round(pf(summary_full$fstatistic[1], summary_full$fstatistic[2], 
                    summary_full$fstatistic[3], lower.tail = FALSE), 4)),
    round(summary_full$sigma, 4),
    paste(summary_full$df[2], "on", summary_full$df[1], "DF")
  )
)

kable(regression_stats)
```

### Koeffizienten mit Konfidenzintervallen

```{r}
#| label: tbl-coefficients
#| tbl-cap: "Regressionskoeffizienten mit 95% Konfidenzintervallen"

coef_table <- data.frame(
  Parameter = c("Intercept (log a)", "Steigung (b)"),
  Schätzung = round(coef(model), 4),
  Standardfehler = round(summary_full$coefficients[,2], 4),
  t_Wert = round(summary_full$coefficients[,3], 3),
  p_Wert = ifelse(summary_full$coefficients[,4] < 0.001, "< 0.001", 
                  round(summary_full$coefficients[,4], 4)),
  CI_2.5 = round(confint(model)[,1], 4),
  CI_97.5 = round(confint(model)[,2], 4)
)

kable(coef_table)
```

### Modelldiagnostik

```{r}
#| label: fig-diagnostics
#| fig-cap: "Diagnostische Plots für die Regressionsanalyse"
#| fig-height: 8

par(mfrow = c(2, 2))
plot(model, which = 1:4, caption = c("Residuen vs. Angepasste Werte", 
                                     "Q-Q Plot der Residuen",
                                     "Scale-Location Plot", 
                                     "Residuen vs. Leverage"))
par(mfrow = c(1, 1))
```

